Machine Learning based TIR predictor

### 2020-09-11 main.tex
Squashed 'sub/SynBio_ML_Writing/' changes from f3cb5d2..19f94fb
19f94fb update ML methods with plots
1d36fa7 fix bugs
9c1aae6 Updated appendix.
60a3ebd move some ML content to method sec.
33d61d0 Updated flowchart
45c8931 Updated methods.tex
c3f267f Update methods.tex
6b82ce7 Updated methods section.
2f52ce4 update batch ucb

git-subtree-dir: sub/SynBio_ML_Writing
git-subtree-split: 19f94fb2cf879238c8fcfd1c1d76d8eb86ce29d7

Our overall experimental goal was to optimise the translation initiation rate by identifying the set of RBS sequences with top TIR scores with minimising the number of DBTL cycle turns that we had to do. We did this by designing a sequential experimental workflow, where we start with either randomised RBS sequences designed to explore the experimental space (round 1) or with RBS sequences recommended by the algorithm (subsequent rounds). The designs were then physically constructed in batches of 90 to fit our automated process (see \textbf{Methods} section). After constructing the plasmids harbouring the new devices are tested in microplate reader and flow cytometer. The results are then fed back to the algorithm for it to recommend the next round of designs.\\

\textcolor{red}{DBTL cartoon}

\subsection{LEARN - prediction of RBS performance}
First problem to be tackled was how to embed the sequence to give our RBS sequences numerical form. As we will be using a Gaussian Process for regression we have investigated use of different types of kernels (also called covariances) for embedding \cite{Ben-Hur2008}. The additional, and positive, effect of using kernels is that our data will be moved to a higher-dimension space, which makes the regression process easier. We compared performance of Dot Product, RBS and a number of string kernels: spectrum, mixed spectrum, weighted degree and weighted degree with shifting \textcolor{red}{Figure X}. Since we found that the spectrum kernel performed the best, we have used it in subsequent studies. More specifically, we used a summary of three kernels: spectrum kernel to process the core 6bp and dot product kernel to process the 7bp flanking sequences both upstream and downstream of the core sequence.\\
The predictions were made using the Gaussian Process regression (GPR) algorithm. In essence, Gaussian Process is a stochastic (Bayesian) predictor of the shape of a function, which is built based on  perceived similarities between data points (kernels), where not only a mean value, but also probability distribution of the mean is calculated. Such an approach makes GPR well suited to predict biological phenomena which are also highly stochastic. \\
\textcolor{red}{Figure X} shows \\

\textcolor{red}{Do we need a figure showing comparison of different kernel's performance? I think it would be useful}

\subsection{DESIGN of the genetic device}
There is a number of factors that impact the protein expression rate, many of them concerned with how the ribosome recognises and binds to the RBS sequence \cite{Chen1994,Vellanoweth1992}. In \emph{E. coli} the RBS is usually located in the 20 bases upstream of the start codon. The RBS usually has a distinguishable, consensus, core sequence called the Shine-Dalgarno sequence, which in \emph{E. coli} is AGGAGG. Here, we put that 20 bp long sequence into focus with main emphasis being put on the 6bp core region (Figure 1).\\
In our genetic design, the investigated RBS controls expression of the Green Fluorescent Protein (GFP) in its mut3b variant. By controlling expression of a fluorescent protein with the RBS we can quickly assess the perceived TIR by measuring fluorescence of cells harbouring plasmid with the device. Finally, the mRNA is transcribed from an IPTG-inducible promoter pLlacO-1. By making the whole device inducible we can synchronise the start of the expression of the GFP in all the cultures by inducing them at the same optical density (OD\textsubscript{600}) with addition of IPTG.\\
The investigated RBS sequence is 20 bps long with the sequence TTTAAGAAGGAGATATACAT, which is a known high TIR RBS that comes with the pBb series plasmids \cite{Lee2011}. In our design we focus on randomising of the core -8 to -13 (relative to the GFP) nucleotides of the RBS and fix others to be the same as the consensus sequence, i.e. TTTAAGA + NNNNNN + TATACAT. Since for each of the 6 position there are 4 possibilities: A, C, G, T the total variant space is $4^6$ = 4096.\\
Since there was no prior data that we could have used to guide our design for the first round of experiments, we have designed 180 (two times 90) RBS sequences based on the consensus sequence that we expected to give good cover of the experimental space:

\begin{enumerate}
    \item 60 RBS sequences which are subsequent single nucleotide changes of all 20 nucleotides of the original, consensus sequence. This batch is designed to show us influence of such single nucleotide changes on the overall performance of the RBS and the potential impact of changes made beyond the core part.
    \item 30 RBS sequences that were fully and uniformly randomised (equal probability of choosing either nucleotide for each position) 
    \item 30 RBS sequences randomised based on the position probability matrix (PPM) \textcolor{red}{we need a citation and better explanation for this}  
    \item 60 RBS sequences recommended by initial machine learning approach. For this, and subsequent design rounds, we used multi-armed bandit recommendation algorithm using predictions made by the Gaussian Process regression (see \textbf{Methods} section) \cite{srinivas2012information}. Since we don't have access to any prior data fitting our design we approximated the predictions using the data from \textcite{jervis2018machine}. This set contains 113 non-repeated records for 56 unique RBS sequences with the respective TIR. The label values are between 0 - 100,000 and skewed, which is shown in Figure xx. First, we have normalised the label to 0 - 1 using the min-max normalisation. Next, the data has been processed using the GPUCB and bandit algorithms as described below.
\end{enumerate}{}

@@

The bandit algorithm aims at maximising the reward (output) from testing a limited number of instances from a big pool which cannot be wholly tested due to limited resources (time, computational power, capital). As such, it is very useful in solving problems like the one presented here - a big pool of potential designs, but only limited time and money that can be used for finding the optimal one.\\
In short, the multi-armed bandit algorithm is a stochastic method of probing of the experimental space. In our case we use the Upper Confidence Bound \textcolor{red}{flavour} of the algorithm, which focuses its recommendations on sequences that should give highest TIR based on the probabilities computer by the prediction algorithm (GPR). Another feature of the bandits algorithms is that it balances two approaches: exploration and exploitation.vExploration makes the algorithm to recommend designs that will improve the predictions better, whereas exploitation will recommend designs that focus on delivering the most efficient design the fastest. The two approaches can be controlled with the \textbeta\enspace parameter. We have decided that in the first iterations of the cycle it would be beneficial to skew the algorithm towards the exploration with exploitation taking increasing role in later iterations. One thing of note is that the bandit algorithm is stochastic, that is it exploits the probabilities of given event occurring (in this case RBS having a specific TIR). As such, it pairs naturally with our prediction algorithm, the Gaussian Process, which provides probability based function regression.\\
\\
For rounds beyond the first initial one, the designs were recommended only by the bandit algorithm based on the data obtained in the previous one, without adding any random designs. In total, three more rounds beyond the initial one have been performed, each with designs recommended by the algorithm.

>>>>

Our overall experimental goal was to maximise the Translation Initiation Rate (TIR) by identifying the set of RBS sequences with top TIR scores while minimising the number of DBTL cycle turns that we had to do. We did this by designing a sequential experimental workflow, where we start with either randomised RBS sequences designed to explore the experimental space (designated round zero since only preexisting, literature data is used for its design) or with RBS sequences recommended by the algorithm (subsequent rounds). The designs were then physically constructed in batches of 90 to fit our automated process (see \textbf{Methods} section). After constructing, the plasmids harbouring the new genetic devices are tested in microplate reader and flow cytometer. The results are then fed back to the algorithm for it to recommend the next round of designs.\\ \textcolor{red}{Figure X} shows the DBTL cycle with the machine learning process emphasised. Our machine learning workflow encompasses two phases of the cycle, namely the LEARN and DESIGN phase. In LEARN phase we use the Gaussian Process algorithm to predict the behaviour of different RBS sequences and in the DESIGN phase we use the multi-armed Bandit algorithm to recommend sequences to be Built and Tested next. Conventionally, the Design phase is considered the beginning of the cycle, however we will start description of our results from the learn phase, which we find a more natural starting point for a machine learning centred workflow.\\

\subsection{LEARN - prediction of RBS performance}
First problem to be tackled was how to embed the sequence to give our RBS sequences numerical form. As we will be using a Gaussian Process for regression we have investigated use of different types of kernels (also called covariances) for embedding \cite{Ben-Hur2008}. The additional, and positive, effect of using kernels is that our data will be moved to a higher-dimensional space, which makes the regression process easier. We compared performance of Dot Product, RBF and a number of string kernels: spectrum, mixed spectrum, weighted degree and weighted degree with shifting \textcolor{red}{Figure Xa}. Since we found that the spectrum kernel performed the best, we have used it in subsequent studies. More specifically, we used a summary of three kernels: spectrum kernel to process the core 6bp and dot product kernel to process the 7bp flanking sequences both upstream and downstream of the core sequence. This approach allowed for good balance between computational complexity and performance.\\
The predictions were made using the Gaussian Process regression (GPR) algorithm. In essence, Gaussian Process is a stochastic (Bayesian) predictor of the shape of a function, which is built based on  perceived similarities between data points (kernels), where not only a mean value, but also probability distribution of the mean is calculated. Such an approach makes GPR well suited to predict biological phenomena which are also highly stochastic. \\
For the zeroth round, since we don't have access to any prior data fitting our design we approximated the predictions using the data from \textcite{jervis2018machine} and the Gaussian Process regression method (see \textbf{Methods} section) \cite{srinivas2012information}. This set contains 113 non-repeated records for 56 unique RBS sequences with the respective TIR. The label values are between 0 - 100,000 and skewed, which is shown in Figure xx. First, we have normalised the label to 0 - 1 using the min-max normalisation. The predictions are shown in \textcolor{red}{Figure Xb}\\
The predictions for next rounds were predicted similarly, but with use of data obtained in previous rounds, visualisation of prediction for the last round of designs is shown in \textcolor{red}{Figure Xc} \\

\subsection{DESIGN of the genetic device}
There is a number of factors that impact the protein expression rate, many of them concerned with how the ribosome recognises and binds to the RBS sequence \cite{Chen1994,Vellanoweth1992}. In \emph{E. coli} the RBS is usually located in the 20 bases upstream of the start codon. The RBS usually has a distinguishable, consensus, core sequence called the Shine-Dalgarno sequence, which in \emph{E. coli} is AGGAGG. Here, we put that 20 bp long sequence into focus with main emphasis being put on the 6bp core region (Figure 1).\\
In our genetic design, the investigated RBS controls expression of the Green Fluorescent Protein (GFP) in its mut3b variant. By controlling expression of a fluorescent protein with the RBS we can quickly assess the perceived TIR by measuring fluorescence of cells harbouring plasmid with the device. Finally, the mRNA is transcribed from an IPTG-inducible promoter pLlacO-1. By making the whole device inducible we can synchronise the start of the expression of the GFP in all the cultures by inducing them at the same optical density (OD\textsubscript{600}) with addition of IPTG.\\
The investigated RBS sequence is 20 bps long with the sequence TTTAAGAAGGAGATATACAT, which is a known high TIR RBS that comes with the pBb series plasmids \cite{Lee2011}. In our design we focus on randomising of the core -8 to -13 (relative to the GFP) nucleotides of the RBS and fix others to be the same as the consensus sequence, i.e. TTTAAGA + NNNNNN + TATACAT. Since for each of the 6 position there are 4 possibilities: A, C, G, T the total variant space is $4^6$ = 4096.\\
Since there was no prior data that we could have used to guide our design for the zeroth round of experiments, we have designed 180 (two times 90) RBS sequences based on the consensus sequence that we expected to give good cover of the experimental space:

\begin{enumerate}
    \item 60 RBS sequences which are subsequent single nucleotide changes of all 20 nucleotides of the original, consensus sequence. This batch is designed to show us influence of such single nucleotide changes on the overall performance of the RBS and the potential impact of changes made beyond the core part.
    \item 30 RBS sequences that were fully and uniformly randomised (equal probability of choosing either nucleotide for each position) 
    \item 30 RBS sequences randomised based on the position probability matrix (PPM) \textcolor{red}{we need a citation and better explanation for this}  
    \item 60 RBS sequences recommended by our multi armed bandit recommendation algorithm based on the initial literature, as described below.
\end{enumerate}{}

@@

The bandit algorithm aims at maximising the reward (output) from testing a limited number of instances from a big pool which cannot be wholly tested due to limited resources (time, computational power, capital). As such, it is very useful in solving problems like the one presented here - a big pool of potential designs, but only limited time and money that can be used for finding the optimal one.\\
In short, the multi-armed bandit algorithm is a stochastic method of probing of the experimental space. In our case we use the Upper Confidence Bound version of the algorithm, which focuses its recommendations on sequences that should give highest TIR based on the probabilities computer by the prediction algorithm (GPR). Another feature of the bandits algorithms is that it balances two approaches: exploration and exploitation. Exploration makes the algorithm to recommend designs that will improve the predictions better, whereas exploitation will recommend designs that focus on delivering the most efficient design the fastest. The two approaches can be controlled with the \textbeta\enspace parameter. We have decided that in the first iterations of the cycle it would be beneficial to skew the algorithm towards the exploration with exploitation taking increasing role in later iterations. One thing of note is that the bandit algorithm is stochastic, that is it exploits the probabilities of given event occurring (in this case RBS having a specific TIR). As such, it pairs naturally with our prediction algorithm, the Gaussian Process, which provides probability based function regression.\\
\\
For rounds beyond the initial, zeroth one, the designs were recommended only by the bandit algorithm based on the data obtained in the previous one, without adding any random designs. In total, three more rounds beyond the initial one have been performed, each with designs recommended by the algorithm.
